\documentclass[a4paper]{llncs}
\usepackage{url}

\newcommand{\kw}[1]{\textsf{#1}}
\newcommand{\caduceus}{\textsc{Caduceus}}
\newcommand{\minelt}[1]{\ensuremath{\mathit{min\_elt}(#1)}}

\begin{document}

\title{Queens on a Chessboard: \\
       an Exercise in Program Verification}
\author{Jean-Christophe Filli\^atre}
\institute{CNRS -- Universit\'e Paris Sud \\ 91405 Orsay, France}
\maketitle

\begin{abstract}
  This article details the formal verification of a 2-lines C program
  which computes the number of solutions to the $N$-queens problem.
\end{abstract}

\section{Introduction}

Formal methods deal either with small properties of big programs or
big properties of small programs. This paper describes a case study
which definitely belongs to the latter category.
Indeed, it details the formal verification of the following 2-lines C program:
{\small\begin{verbatim}
t(a,b,c){int d=0,e=a&~b&~c,f=1;if(a)for(f=0;d=(e-=d)&-e;f+=t(a-d,(b+d)*2,(
c+d)/2));return f;}main(q){scanf("%d",&q);printf("%d\n",t(~(~0<<q),0,0));}
\end{verbatim}}
\noindent 
This rather obfuscated code was found on a web page gathering C signature
programs\footnote{\url{http://www.iwriteiam.nl/SigProgC.html}} and is
apparently due to Marcel van Kervinc. This is a standalone C program
which reads an integer on the standard input, say $n$, and then prints another
integer on the standard output, say $f(n)$. If $n$ is smaller than the
machine word size (typically 32), then
$f(n)$ appears to be the number of solutions to the famous $n$-queens
problem, that is the number of ways to put $n$ queens on a $n\times n$
chessboard so that they cannot beat each other. And even more
surprisingly, this is a very efficient program to compute this number.

As a case study for the \caduceus\ tool we are
co-developing~\cite{caduceus}, we considered 
the formal verification of this program. By verification, we mean here
the mechanically-assisted proof that this program (1) does not crash, 
(2) terminates, and (3) indeed computes the right number. We find this
challenging since these two lines of code are at the same time idiomatic C
programming, complex from an algorithmic point of view and highly
efficient. Thus we can demonstrate the ability of our tools to tackle
a wide range of verification issues without considering a too long
program. 

This paper is organized as follows. First Section~\ref{unobf}
``unobfuscates'' the program, unravelling the mystery of its algorithm
and data. Then Section~\ref{caduceus} briefly introduces the \caduceus\
tool. Finally Section~\ref{verif} details the formal verification,
namely the logical annotations inserted in the program
and the methods used to discharge the resulting verification conditions.

\section{Unobfuscation}\label{unobf}

Before entering the formal verification process, we first unravel the
mystery behind this obfuscated code. The code is composed of two
parts: a recursive function \texttt{t}, which takes three integers as
arguments and returns an integer; and a main function which reads an
integer from the standard input, makes a call to the function
\texttt{t} and prints the result on the standard output.
With type declarations and a bit of indentation, the function
\texttt{t} reads as follows:
\begin{verbatim}
int t(int a, int b, int c) {
  int d=0,e=a&~b&~c,f=1;
  if(a) for(f=0;d=(e-=d)&-e;f+=t(a-d,(b+d)*2,(c+d)/2));
  return f;
}
\end{verbatim}
The assignment \verb!d=(e-=d)&-e! actually does not conform with the
ANSI norm, because it assumes that the inner assignment \verb!e-=d! is
performed before the evaluation of \verb!-e!. This is not guaranteed
and the compiler may freely choose between the two possible evaluation
strategies. Since the \caduceus\ tool only deals with ANSI C programs,
we need to modify the code. Fortunately, this is quite easy:
since \verb!d! is initialized to 0 we can move the
assignment \verb!e-=d! to the end of the loop body and save the
initialization. We have even reduced the size of the original code.



INPUT/OUTPUT REMOVED AND MAIN REPLACED BY A INT TO INT FUNCTION

We end up with the code given in Figure~\ref{fig:code}.
\begin{figure}[t]
  \centering\hrulefill\vspace{-1em}
\begin{verbatim}
int t(int a, int b, int c) {
  int d, e=a&~b&~c, f=1;
  if (a)
    for (f=0; d=e&-e; e-=d)
      f += t(a-d,(b+d)*2,(c+d)/2));
  return f;
}
int queens(int q) {
  return t(~(~0<<q),0,0);
}
\end{verbatim}  
\vspace{-1.2em}\hrulefill\vspace{-1em}
  \caption{The C code to be verified}
  \label{fig:code}
\end{figure}

INTEGERS USED AS SETS (EXPLAIN); CODE REPHRASED ON SETS READS AS FOLLOWS:

{\begin{obeylines}
  \kw{int} $t$($a$, $b$, $c$) 
  ~~ \kw{if} $a \not= \emptyset$ 
  ~~~~ $e$ $\leftarrow$ $(a \backslash  b) \backslash c$ 
  ~~~~ $f$ $\leftarrow$ 0 
  ~~~~ \kw{while} $e \not=\emptyset$ 
  ~~~~~~ $d$ $\leftarrow$ $\minelt{e}$ 
  ~~~~~~ $f$ $\leftarrow$ $f$ $+$ $t$($a\backslash \{d\}$, $S(b\cup\{d\})$, $P(c\cup\{d\})$) 
  ~~~~~~ $e$ $\leftarrow$ $e \backslash  \{d\}$ 
  ~~~~ \kw{return} $f$ 
  ~~ \kw{else} 
  ~~~~ \kw{return} $1$ 
  ~~ 
  \kw{int} $f$($n$) 
  ~~ \kw{return} $t$($\{0,1,\dots,n-1\}$, $\emptyset$, $\emptyset$)
\end{obeylines}}


\section{The Caduceus tool}\label{caduceus}

\cite{caduceus,FilliatreMarche04} \cite{why}

\cite{simplify} \cite{yices} \cite{ergo} \cite{coq}


\section{Formal verification}\label{verif}

\subsection{Termination}

\subsection{Correctness}

\section{Conclusion}

\nocite{*}
\bibliographystyle{plain}
\bibliography{./biblio}

\end{document}
