\documentclass{llncs}

\usepackage{amssymb}
\usepackage{alltt}
\usepackage{url}
\usepackage{pstricks}
\usepackage{pst-node}
\usepackage{verbatim}
%\usepackage{pslatex}
\usepackage{times}
%\usepackage{listings}

\input{./macros.tex}

% LLNCS hack: too many space around subsection titles
\makeatletter
\def\subsection{\@startsection{subsection}{2}{\z@}{-8pt plus-3pt minus
 -3pt}{5pt plus 3pt minus 0pt}{\normalsize\bf\boldmath
\pretolerance=10000\relax\rightskip=0pt plus8em}}
\makeatother

\pagestyle{plain}

\begin{document}

\title{A case study of C source code verification: the Schorr-Waite algorithm}
\author{Thierry Hubert and Claude March\'e}
\institute{PCRI --- LRI (CNRS UMR 8623) --- INRIA Futurs ---
  Universit\'e Paris 11 \\
B\^at 490, Universit\'e Paris-sud, 91405 Orsay cedex, France \\
\{\texttt{hubert},\texttt{marche}\}\texttt{@lri.fr} 
}
\maketitle

\begin{abstract}
  We describe an experiment of formal verification of C source code,
  using the \caduceus{} tool. We performed a full formal proof of the
  classical Schorr-Waite graph-marking algorithm, which has already
  been used several times as a case study for formal reasoning on
  pointer programs. Our study is original with respect to previous
  experiments for several reasons. First, we use a general-purpose
  tool for C programs: we start from a real source code written in
  C, specified using an annotation language for arbitrary C programs.
  Second, we use several theorem provers as back-ends, both automatic
  and interactive.  Third, we indeed formally establish more
  properties of the algorithm than previous works, in particular a
  formal proof of termination is made.

  \noindent{\bf Keywords:} Formal verification, Floyd-Hoare logic,
  Pointer programs, Aliasing, C programming language.
\end{abstract}

\footnotetext{This research was partly supported by the ``projet
  GECCOO de l'ACI S\'ecurit\'e Informatique'',
  \url{http://gecco.lri.fr}, the AVERROES RNTL (national network on
  software technology) project
  \url{http://www-verimag.imag.fr/AVERROES/}, CNRS \& INRIA,}

\input{./intro.tex}
\input{./overview.tex}
\input{./algo.tex}
\input{./proof.tex}
\input{./concl.tex}

\bigskip
\paragraph{Acknowledgments.} 
We thank Jean-Christophe Filli\^atre for its fruitful help, advices and
remarks about this case study.

\bibliographystyle{abbrv}
\input{biblio-demons}


\end{document}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: t
%%% End: 
